---
title: "data.table demo"
author: "Brian High"
date: "1/30/2015"
output: 
    html_document: 
        keep_md: yes
---

## Load Packages and Data

Load packages.

```{r, echo=TRUE}
library(data.table)
library(ggplot2)
```

Load the `iris` data.

```{r, echo=TRUE}
data(iris)
head(iris)
```

Convert to a `data.table`.

```{r, echo=TRUE}
irisdt <- as.data.table(iris)
head(irisdt)
```

## Prepare for Reshaping

Add a column for the flower id.

```{r, echo=TRUE}
irisdt[, flower_id := as.numeric(rownames(iris))]
head(irisdt)
```

## Reshape with `melt` and `dcast`

Convert wide data format to long format.

```{r, echo=TRUE}
irisdt_melted <- melt(irisdt, 
                      id = c("flower_id", "Species"), 
                      measure = c("Sepal.Length", "Sepal.Width", 
                                  "Petal.Length", "Petal.Width"))
head(irisdt_melted)
```

Add new columns for the parsed values. Remove the variable column.

```{r, echo=TRUE}
irisdt_melted[, flower_part := gsub("(\\w*)\\.\\w*", "\\1", variable)]
irisdt_melted[, measurement_type := gsub("\\w*\\.(\\w*)", "\\1", variable)]
irisdt_melted[, variable := NULL]
head(irisdt_melted)
```

Convert `measurement_type` to columns in wide format.

```{r, echo=TRUE}
irisdt_dcast <- dcast(irisdt_melted, 
                     formula=flower_id+Species+flower_part~measurement_type)
head(irisdt_dcast)
```

## Plot with `qplot`

Produce faceted plot with ggplot2's `qplot`.

```{r, echo=TRUE}
qplot(x=Width, y=Length, data=irisdt_dcast, geom=c("point","smooth"), 
      color=Species, method="lm", facets=flower_part~Species)
```

## Plot with `ggplot`

Produce a faceted plot with ggplot2's `ggplot` instead of `qplot`.

```{r, echo=TRUE}
ggplot(data=irisdt_dcast, aes(x=Width, y=Length)) + 
    geom_point() + facet_grid(Species~flower_part, scale="free") +
    geom_smooth(method="lm") + theme_bw(base_size=16)
```
