---
title: "dplyr and tidyr demonstration"
author: "Brian High"
date: "Jan. 18th, 2015"
output:
     ioslides_presentation:
         fig_caption: yes
         fig_retina: 1
         fig_width: 6.5
         keep_md: yes
         smaller: yes
---

## Introduction
This example shows an alternate way to generate the plot from Raphael
Gottardo's RMarkdown presentation: 
[Advanced_graphics_in_R.Rmd](https://github.com/raphg/Biostat-578/blob/master/Advanced_graphics_in_R.Rmd)  

The code presented here was written by Brian High, except for the `qplot`
and `ggplot` plotting code, which was modified from Raphael Gottardo's original.

To the extent possible under law, Brian High has waived all copyright and 
related or neighboring rights to "dplyr and tidyr demonstration". This work 
is published from: United States. 

## Load packages and data
Load ggplot2 and iris data.

```{r, echo=TRUE}
suppressMessages(library(ggplot2))
data(iris)
```

Load dplyr and tidyr.

```{r, echo=TRUE}
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
```

## dplyr: `mutate`
Using `mutate`, add a column to keep track of the flower id.

```{r, echo=TRUE}
iris_id <- mutate(iris, flower_id = rownames(iris))
head(iris_id)
```

## tidyr: `gather`
Convert wide data format to long format.

```{r, echo=TRUE}
iris_gathered <- gather(iris_id, variable, value, c(-Species, -flower_id))
head(iris_gathered)
```

## dplyr: `mutate` and `gsub`
Add new columns for the parsed values, remove the variable column.

```{r, echo=TRUE}
iris_parsed <- mutate(iris_gathered, 
                      flower_part = gsub("(\\w*)\\.\\w*", "\\1", variable), 
                      measurement_type = gsub("\\w*\\.(\\w*)", "\\1", variable),
                      variable = NULL)
head(iris_parsed)
```

## tidyr: `extract`
You can also do the parsing and column creation with tidyr's `extract`.

```{r, echo=TRUE}
iris_parsed <- extract(iris_gathered, variable, 
                       c("flower_part", "measurement_type"), "(.+)\\.(.+)")
    
head(iris_parsed)
```

## tidyr: `spread`
Convert measurement_types to columns in wide format.

```{r, echo=TRUE}
iris_spread <- spread(iris_parsed, measurement_type, value)
head(iris_spread)
```

## Plot with ggplot2's `qplot`
Produce faceted plot with ggplot2's `qplot`.

```{r, echo=TRUE, fig.height=4}
qplot(x=Width, y=Length, data=iris_spread, geom=c("point","smooth"), 
      color=Species, method="lm", facets= flower_part~Species)
```


## Repeat using a pipe
All of the data tidying could be done in one "pipe line".

```{r, echo=TRUE}
iris_spread <- mutate(iris, flower_id = rownames(iris)) %>%
    gather(variable, value, c(-Species, -flower_id)) %>%
    extract(variable, c("flower_part", "measurement_type"), "(.+)\\.(.+)") %>%
    spread(measurement_type, value)
```

## Plot with ggplot2's `qplot` again
Produce the faceted plot again with ggplot2's `qplot`.

```{r, echo=TRUE, fig.height=4}
qplot(x=Width, y=Length, data=iris_spread, geom=c("point","smooth"), 
      color=Species, method="lm", facets= flower_part~Species)
```

## Plot with `ggplot`
Produce a faceted plot with ggplot2's `ggplot` instead of `qplot`.

```{r, echo=TRUE, fig.height=4}
ggplot(data=iris_spread, aes(x=Width, y=Length)) + 
    geom_point() + facet_grid(Species~flower_part, scale="free") +
    geom_smooth(method="lm") + theme_bw(base_size=16)
```

