---
title: 'Bioinformatics for Big Omics Data: Tidyr Example'
author: "Raphael Gottardo and Brian High"
date: "December 13, 2014"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    keep_md: yes
    smaller: yes
---

## Let's load ggplot2

```{r, excho=TRUE}
library(ggplot2)
```

## Reshaping your data with reshape2

It is often necessary to reshape (e.g. pivot) your data before analysis. This can easily be done in R using the `reshape2` package.
This package provides main functions `melt` and `*cast`. `melt` basically "melts" a dataframe in wide format into a long format. `*cast` goes in the other direction.

## Reshaping your data with reshape2

Let's revisite our `iris` dataset.
```{r reshape2}
# We first load the library
library(reshape2)
# Only display the first few lines
head(iris)
```

We can see in the data above that we are measuring both width and length on two different flower characteristics: Sepal, and Petal. So we could store the same information with only one length (resp. width) column and an additional variable for type (Sepal/Petal). 

## reshape2 - melt

The `melt` function provides some good default options that will try to best guess how to "melt" the data.

```{r reshape2-melt}
# We first need to add a column to keep track of the flower
iris$flower_id <- rownames(iris)
# Default options
iris_melted <- melt(iris)
head(iris_melted)
```

## tidyr - gather

Or, alternatively, using tidyr and "gather" instead of reshape2 and "melt":

```{r, tidyr-gather}
library(tidyr)
iris_gathered <- gather(iris, variable, value, c(-Species, -flower_id))
head(iris_gathered)
```

## reshape2 - melt

```{r reshape2-melt-suite}
# We first split that variable to get the columns we need
split_variable <- strsplit(as.character(iris_melted$variable),split="\\.")
# Create two new variables
iris_melted$flower_part <- sapply(split_variable, "[", 1)
iris_melted$measurement_type <- sapply(split_variable, "[", 2)
# Remove the one we don't need anymore
iris_melted$variable <- NULL
head(iris_melted)
```

This is close but not quite what we want, let's see if cast can help us do what we need.

## tidyr - gather

And the same operation, but instead with the "gathered" data.frame...

```{r tidyr-gather-suite}
# We first split that variable to get the columns we need
split_variable <- strsplit(as.character(iris_gathered$variable),split="\\.")
# Create two new variables
iris_gathered$flower_part <- sapply(split_variable, "[", 1)
iris_gathered$measurement_type <- sapply(split_variable, "[", 2)
# Remove the one we don't need anymore
iris_gathered$variable <- NULL
head(iris_gathered)
```

## reshape2 - cast

Use `acast` or `dcast` depending on whether you want vector/matrix/array output or data frame output. Data frames can have at most two dimensions.

```{r reshape2-cast}
iris_cast <- dcast(iris_melted, formula=flower_id+Species+flower_part~measurement_type)
#head(iris_cast)
```

**Q:** Why are the elements of `flower_id` not properly ordered?

`melt` and `*cast` are very powerful. These can also be used on `data.tables`. More on this latter.

**Exercise:** Try to reorder the variable names in the formula. What happens?

## tidyr - spread

And the same plot using spread with the tidyr "gathered" data.frame...

```{r tidyr-spread}
iris_spread <- spread(iris_gathered, measurement_type, value)
head(iris_spread)
```

## Back to ggplot2

Using our long format dataframe, we can further explore the iris dataset.

```{r multi-facet, fig.height=4}
# We can now facet by Species and Petal/Sepal
qplot(x=Width, y=Length, data=iris_cast, geom=c("point","smooth"), color=Species, method="lm", facets= flower_part~Species)
```

It would be nice to see if we could have free scales for the panels, but before we explore this, let's talk about the `ggplot` API as an alternative to qplot. Can we also customize the look and feel?

## Back to ggplot2 - tidyr

And again, using the tidyr data.frame...

```{r multi-facet-tidyr, fig.height=4}
qplot(x=Width, y=Length, data=iris_spread, geom=c("point","smooth"), color=Species, method="lm", facets= flower_part~Species)
```
